Isolator: accurate and consistent analysis of isoform-level expression in RNA-Seq experiments

### 2016-02-24 supplement.tex
Add Nathan's story. Misc improvements.
Supplement work.

\documentclass{article}
\usepackage[T1]{fontenc}
\usepackage[cm-default]{fontspec}
\usepackage{lmodern}

@@

% TODO:
%   * Explain that every SEQC sample was put in the same experiment for
%     isolator.

@@

\begin{table}
\begin{center}
\begin{tabular}[c]{@{}lrr@{}}
\toprule\addlinespace
Method & $c$ vs $0.75a + 0.25b$ & $d$ vs $0.25a + 0.75b$
\\\addlinespace
\midrule
Isolator & \textbf{0.995} & \textbf{0.995}
\\\addlinespace
BitSeq & 0.994 & 0.994
\\\addlinespace
RSEM/PM & 0.994 & 0.993
\\\addlinespace
RSEM/ML & 0.987 & 0.987
\\\addlinespace
Salmon & 0.986 & 0.986
\\\addlinespace
eXpress & 0.983 & 0.983
\\\addlinespace
Sailfish & 0.978 & 0.976
\\\addlinespace
Kallisto & 0.972 & 0.970
\\\addlinespace
Cufflinks & 0.941 & 0.982
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between gene-level estimates for
the mixed samples C and D and weighted averages of estimates for A and
B, corresponding to the mixture proportions for C and D. TODO: move this
table to the supplement.}
\end{center}
\label{table:seqcgenes}
\end{table}

@@

>>>>

\documentclass{article}
\usepackage[left=1in, right=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[cm-default]{fontspec}
\usepackage{lmodern}

@@

% TODO:
%   * Explain that every SEQC sample was put in the same experiment for
%     isolator.
%
%   * Explain the reasoning behind adding 0.1 tpm to expression values to
%     account for zeros.

@@

\begin{table}
\begin{center}
\begin{tabular}[c]{@{}lrr@{}}
\toprule\addlinespace
Method & $c$ vs $0.75a + 0.25b$ & $d$ vs $0.25a + 0.75b$
\\\addlinespace
\midrule
Isolator & \textbf{0.995} & \textbf{0.995}
\\\addlinespace
BitSeq & 0.994 & 0.994
\\\addlinespace
RSEM/PM & 0.994 & 0.993
\\\addlinespace
RSEM/ML & 0.987 & 0.987
\\\addlinespace
Salmon & 0.986 & 0.986
\\\addlinespace
eXpress & 0.983 & 0.983
\\\addlinespace
Sailfish & 0.978 & 0.976
\\\addlinespace
Kallisto & 0.972 & 0.970
\\\addlinespace
Cufflinks & 0.941 & 0.982
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between gene-level estimates for
the mixed samples C and D and weighted averages of estimates for A and
B, corresponding to the mixture proportions for C and D.}
\end{center}
\label{table:seqcgenes}
\end{table}

@@

\section{Notes on measuring agreement between transcript expression estimates}

Benchmarks in this paper rely heavily on quantifying the agreement between
expression estimates and qPCR, spike-in controls, estimates from other samples,
or ground truth (in the case of simulations). As discussed in Section
\ref{results}, our methodology uses proportionality correlation as an
alternative to more common measures of correlation or aggregate error.

In practice, the distribution of transcript expression values tends to be skewed
right with a very low median. This can lead to misleading comparisons in two
ways. First, similarity/dissimilarity measurements like Pearson's correlation or
root mean square error may be dominated by a small number of very highly
expressed transcripts. Second, and conversely, measurements like Spearman's rank
correlation may be dominated by unexpressed, or very low-expression transcripts
which often make up the large majority of transcripts in a relatively
well-annotated species like human.

This problem is compounded by the use of different estimators in the methods
being compared.  When maximum likelihood or, depending on the choice of prior,
maximum posterior estimates are computed, many expression estimates will be
zero. On the other hand, posterior mean estimates (as is computed in Isolator,
BitSeq, and optionally RSEM) are inherently non-zero.

In deciding how to deal with zeros, or the lack of zeros, it is easy to
artificially advantage one approach or the other. For example, removing zeros,
disadvantages methods that have no zeros, whereas quantifying error with
functions like median relative difference (as was used to argue the accuracy of
Kallisto \cite{Bray:2015uj}), will disadvantage methods that do not produce
zeros, or otherwise report small numbers.
\footnote{
To see this, consider that the absolute relative difference
between 0 and $\epsilon$ for an arbitrarily large or arbitrarily small but
positive $\epsilon$ is $\frac{\epsilon - 0}{(\epsilon + 0)/2} = 2$. Thus, MRD
considers ``absent versus barely present'' to be equivalent to ``absent versus
highly expressed'', an outcome that is at best, unintuitive.

More generally, the scale-invariance of relative difference means

%% TODO: which benchmark should we use to explore these problems? Maybe the
% SEQC transcript "consistency" benchmark? Regardless, it seems I didn't save
% any of this data locally so I have to spin up an instance.
% Fuck, I've paid about $600 in fees. I need to stop this thing.

$\frac{\alpha |x_1 - x_2|}{\alpha()/2} = 2$

}

This presents two plausible strategies to compensate: truncate small values to
zero, or add a small constant to every value. Both

\subsection{The effects of truncation}

%This paper argues that posterior mean estimates have some practical benefits.
%And reporting non-zero estimates in all cases is justifiable from a Bayesian
%...

\subsection{Effects of varying additive constant on computed proportionality
            correlation}

As our analysis relies on adding a seemingly arbitrary constant $c$ to
expression values before computing correlation, a natural question is how
sensitive our results our to its precise value. In short, we believe it is quite
insensitive, with results largely consistent when $c$ is varied across several
orders of magnitude.

%% TODO: plot

Given the insensitivity, we claim that $c = 0.1$ reasonable choice. If we assume
between $10^5$ to $10^6$ mRNA are present in a mammalian cell
\cite{Islam:2014is}, then adding $0.1 \text{TPM}$ can be interpreted as
negligibly inflating expression estimates by 0.01 to 0.1 copies per cell.

\subsection{Select analyses repeated with Spearman's rank correlation}

While we believe proportionality correlation is a more meaningful measure, our
results remain qualitatively similar using Spearman's rank correlation. In this
section we reproduce tables from the main paper using this somewhat more
conventional approach.

\begin{figure*}
\begin{subtable}{0.5\linewidth}
\caption{Comparison of estimates to TaqMan qPCR}
\begin{tabular}[c]{@{}lrrrr@{}}
\toprule\addlinespace
Method & A & B & C & D
\\\addlinespace
\midrule
Isolator & \textbf{0.903} & \textbf{0.892} & \textbf{0.878} & \textbf{0.859}
\\\addlinespace
eXpress & 0.897 & 0.885 & 0.869 & 0.849
\\\addlinespace
Cufflinks & 0.897 & 0.886 & 0.869 & 0.849
\\\addlinespace
Salmon  &  0.895 & 0.881 & 0.868 & 0.845
\\\addlinespace
RSEM/ML  & 0.895 & 0.880 & 0.867 & 0.844
\\\addlinespace
Kallisto &  0.890 & 0.873 & 0.862 & 0.837
\\\addlinespace
BitSeq  &  0.886 & 0.874 & 0.862 & 0.842
\\\addlinespace
Sailfish &  0.875 & 0.843 & 0.842 & 0.804
\\\addlinespace
RSEM/PM  & 0.886 & 0.873 & 0.862 & 0.840
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\label{table:taqman:spearman}
\end{subtable}
\begin{subtable}{0.5\linewidth}
\caption{Comparison of estimates to PrimePCR qPCR}
\begin{tabular}[c]{@{}lrrrr@{}}
\toprule\addlinespace
Method & A & B & C & D
\\\addlinespace
\midrule
Isolator  & \textbf{0.872} & \textbf{0.863} & \textbf{0.837} & \textbf{0.844}
\\\addlinespace
Cufflinks & 0.864 & 0.855 & 0.826 & 0.835
\\\addlinespace
eXpress   & 0.864 & 0.853 & 0.825 & 0.833
\\\addlinespace
Salmon    & 0.858 & 0.846 & 0.821 & 0.825
\\\addlinespace
RSEM/ML   & 0.856 & 0.845 & 0.819 & 0.823
\\\addlinespace
BitSeq    & 0.855 & 0.842 & 0.821 & 0.827
\\\addlinespace
Kallisto  & 0.855 & 0.840 & 0.817 & 0.821
\\\addlinespace
Sailfish  & 0.840 & 0.811 & 0.797 & 0.793
\\\addlinespace
RSEM/PM   & 0.850 & 0.838 & 0.815 & 0.819
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\label{table:primepcr:spearman}
\end{subtable}

\begin{subtable}{0.5\linewidth}
\caption{Comparison of estimates to ERCC spike-in controls}
\begin{tabular}[c]{@{}lrrrr@{}}
\toprule\addlinespace
Method & A & B & C & D
\\\addlinespace
\midrule
Isolator  & \textbf{0.974} & \textbf{0.973} & \textbf{0.975} & \textbf{0.976}
\\\addlinespace
Salmon    & 0.969 & 0.968 & 0.966 & 0.966
\\\addlinespace
Kallisto  & 0.964 & 0.964 & 0.961 & 0.965
\\\addlinespace
Sailfish  & 0.968 & 0.961 & 0.959 & 0.959
\\\addlinespace
Cufflinks & 0.964 & 0.964 & 0.961 & 0.964
\\\addlinespace
RSEM/PM   & 0.950 & 0.946 & 0.951 & 0.948
\\\addlinespace
RSEM/ML   & 0.945 & 0.950 & 0.948 & 0.948
\\\addlinespace
BitSeq    & 0.941 & 0.938 & 0.945 & 0.942
\\\addlinespace
eXpress   & 0.937 & 0.944 & 0.941 & 0.942
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\label{table:ercc:spearman}
\end{subtable}
\begin{subtable}{0.5\linewidth}
\caption{Consistency of SEQC sample estimates}
\begin{tabular}[c]{@{}lrr@{}}
\toprule\addlinespace
Method & $c$ vs $0.75a + 0.25b$ & $d$ vs $0.25a + 0.75b$
\\\addlinespace
\midrule
Isolator  & \textbf{0.972} & \textbf{0.972}
\\\addlinespace
BitSeq    & 0.946 & 0.947
\\\addlinespace
RSEM/PM   & 0.936 & 0.935
\\\addlinespace
Sailfish  & 0.882 & 0.877
\\\addlinespace
RSEM/ML   & 0.841 & 0.838
\\\addlinespace
Salmon    & 0.823 & 0.821
\\\addlinespace
Kallisto  & 0.845 & 0.842
\\\addlinespace
eXpress   & 0.868 & 0.867
\\\addlinespace
Cufflinks & 0.865 & 0.864
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\label{table:seqctranscripts:spearman}
\end{subtable}

\caption{
Tables from Figure 1 recomputed using Spearman's rank correlation.
}
\end{figure*}

\begin{table}
\begin{center}
\label{table:rlsim:spearman}
\begin{tabular}[c]{@{}ll@{}}
\toprule\addlinespace
Method & Correlation
\\\addlinespace
\midrule
Isolator   & 0.724
\\\addlinespace
Kallisto   & 0.727
\\\addlinespace
Salmon     & \textbf{0.728}
\\\addlinespace
RSEM/ML    & 0.712
\\\addlinespace
Cufflinks  & 0.636
\\\addlinespace
eXpress    & 0.605
\\\addlinespace
Sailfish   & 0.648
\\\addlinespace
RSEM/PM    & 0.605
\\\addlinespace
BitSeq     & 0.595
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Table 1 recomputed using Spearman's rank correlation.}
\end{center}
\end{table}
